Mutual Modeling in Human-Robot Interaction - Scientific part

### 2014-03-20 project-mutual-modeling.tex
Mostly better latex formatting

In order to build a shared understanding, do partners have to build a
representation of each other's knowledge? We refer to \emph{mutual modeling} (MM) as
the process of inferring one's partner mental states. Any claim that students
carry out a detailed monitoring of their peers would be as incorrect as any
claim that they do not maintain any representation at all. If mutual modeling
had to be permanently detailed and accurate, subjects would obviously face a
huge cognitive load. Conversely, peers could not collaborate without some
minimal amount of mutual modeling. For instance, A cannot disagree with B
without knowing that B has a different opinion. The mutual model can be implicit
(A is not aware of what he knows about B), by-default (I believe that B beliefs
what I believe unless contrary evidence), opportunistic (A does not model B
unless the conversation requires it), global (A infers B's beliefs based on
categories such as age, culture or profession) and, of course, it can be
incorrect… but it can not remain empty. Dialogues include many instances of
utterances such as "I thought he would do that" (first level of MM) or even "He
thought I would do that but I intended something else." (second level of MM).

@@

We have different levels of MM. If A states "B thinks I am god in maths", we
are the second level of MM:  $Model(A, B, Model(B,A, knowledge="good in
math"))$. There is an infinite regress of nested models: If A says "B knows that
I don't expect him to solve this statistics problem" corresponds to $Model(A, B,
Model (B, A (Model (A,B, statistic-skills)))$.

A partner model is probably not a "box", \ie not a monolithical representation
but rather a mosaic of information fragments about the partner, with various
grain sizes and various life cycles. This mosaic is elaborated through various
mechanisms, first for building an initial model of the partner and then for
updating this model.  As two students meet for the first time, mutual models are
initialized by the assumptions they make upon each other from cues such as
his/her membership to large categories (age, culture, profession, ...) include
stereotypes (sportsmen, junkie, business women, Swiss,...) as well as physical
appearance. Scholars studied how initial modeling impacts communication. In
their experiments on initial MM, Slugoski, Lalljee, Lamb \&
Ginsburg~\cite{slugoski1993attribution} pretended to their subjects that their
(fake) partner had or had not received the same information. They observed that
the subjects adapted their dialogue by focusing the explanation on the items
that (s)he was supposed to ignore. Brennan (1991) showed that the subjects used
different initial strategies in forming queries depending on who they were told
their partner was.

@@

Mutual modeling does not occur in a vacuum but it is highly contextualized. REF
reviews the features of the collaborative situation, namely the media
(cotemporality… ), may facilitate or hamper mutual modeling. Hutchins (SILENCE)
reported a study in which a short silence between two pilots was perfectly
interpreted because it occurred in a highly constrained communication context.
Some environments are more productive than others in helping peers to detect
their misunderstandings. Roschelle~\cite{roschelle1995construction} reformulate
the design of CSCL interfaces as providing ways for peer to detect and repairs
their misunderstanding. In CSCW, researchers have explored various so-called
"awareness tools", \ie functionalities that inform A is about B's actions that A
could not directly perceived because B was working on a different subset of the
virtual space. Different awareness tools will be addressed in this contribution
since they allowed us to manipulate experimentally the MM activity.

@@

\subsection{Related research of the requirers}

Our knowledge base implements such a mechanism: when the robots infers a new
agent has been introduced in the knowledge base, it initializes a new,
independent, ontology for this agent. All the ontologies that are created share
the same common-sense knowledge, but rely on each agent's perspective for the
actual instantiation: the robot (geometrically) computes that the book is in
its own field of view, but not in the human one. The robot knowledge contains
the fact \stmt{book isVisible true} while the human model contains \stmt{book
isVisible false}.

Using separate knowledge models in the knowledge base, we have been able to
replicate the \emph{Sally and Ann} false-belief experiment with our
robots~\cite{warnier2012when}.

In our implementation, perspective taking is tightly connected to the symbolic
knowledge models, and since our knowledge base allows for storage of one
knowledge model per agent, we have been able to endow the robot with a simple
theory of mind: we explicitly model
what the robot knows about its partners in a symbolic way. This knowledge is
then re-used in different places, to correctly interpret what the human says,
or to plan tasks that are actually doable for the human.

\subsection{Detailed research plan}

\subsubsection{Research questions}

The core research question we want to address in this project is
$(\mathcal{Q}_1)$ how mutual modeling impacts the quality of the human-robot
interaction?  \emph{Quality}, in our context, is measured as the performance of
the human-robot pair in selected joint tasks.

To actually instigate a mutual modeling situation, three other research
questions must be addressed. $(\mathcal{Q}_2)$ how can the robot acquire and
represent a mental model of its partners? $(\mathcal{Q}_3)$ conversely, how do
the human partners model the robot and what are the robot's design/behavioural
factors that influences it? $(\mathcal{Q}_4)$ once the robot detects
misunderstandings, what should be the interaction strategy?
\subsubsection{A theoretical model of mutual modeling}

To report experiments on MM, we use a simple notational system: ``A knows that B
knows X", noted \model{A}{B}{X}. This is a simple but hopefully useful reduction to
reason on mutual modeling. This notation does not mean A has an explicit,
monolithic representation of B: it must be understood as an abstraction referring to complex
socio-cognitive processes.

We refer to the degree of accuracy of the mutual model as
\Model{A}{B}{X}. To estimate the MM effort, we need 3 variables.

@@

    \item Tasks vary a lot with respect to how much they require.  The grounding
        criterion – denoted $\mathcal{M}^{\circ}_{min}$  - refers to how
        important it is to mutually share a piece of information X to succeed
        the task T.. It can be computed as the probability to succeed T despite
        the fact X is not grounded. In the next studies, we estimate
        $\mathcal{M}^{\circ}_{min}(A,B,X)$ as the correlation between
        \Model{A}{B}{X} and task performance.

    \item Before any specific grounding act, there is rarely a null probability
        that X is mutually understood by A and B: because X is part of A's and
        B's cultures, because it is manifest to co-present subjects or simply
        because there is not much space for misunderstanding or disagreement
        about X. We could not simply collaborate without a high level of initial
        grounds. We note the theoretical accuracy of initial grounds:
        $\mathcal{M}^{\circ}_{t1}(A,B,X)$.

    \item The cost of grounding X refers to the physical and cognitive effort
        necessary to perform a grounding act α: a verbal repair (\eg
        rephrasing), a deictic gesture, a physical move to adopt one partner's
        viewpoint, etc. This cost varies according to media features ( Clark \&
        Brennan (REF)).

@@

Based on these 3 parameters, the probability of making an action $\alpha_{t1}$ about
content X at time $t1$ during task T for increasing \Model{A}{B}{X}
at time $t2$ is the ratio between how much it is needed  ((2)-(1)) and how much it
costs (3)(Traum \& Dillenbourg, 1996):

$p(\alpha_{t1}(X,T)   M^{\circ}_{t2}(A,B,X)) ≈ (M^{\circ}_{min}(A,B,X,T) -
M^{\circ}_{t1}(A,B,X))/cost (\alpha)$

This is qualitative summary more that a real equation since several parameters
are hard to quantify (\eg the cost an communication acts depends upon the user
as well) but it clarifies the parameters of the experiments we report hereafter.

@@

    \item $\mathcal{M}^{\circ}_{t1}(A,B,X)$ Along the same reasoning, the initial degree
        of common grounds should be rather low (and hence the difference between
        initial and required degrees rather high) in order to make mutual
        modelling effort more observable. Studies 1, 4 and 5 have been conducted
        with teams of students who did not know each other. They came
        nonetheless from the same university: they hence had some general common
        grounds.  For studies 2 and 3, students knew each other before for
        reasons explained later on. In study 5, we manipulated the initial MM by
        using a JIGSAW script.

    \item $cost(\alpha)$ In all studies but study 4, the cost of  grounding is
        an independent variable. Study 3 uses media richness as independent
        variable, with the hypothesis that modeling emotions is "cheaper" with a
        richer medium, \ie  when peers can see each other.  Studies 1,2 and 4
        use awareness tools which, in principle, reduce the cost of MM, but do
        not eliminate all costs: if the tool provide A with information about
        what B does/knows, this additional information may actually increase
        cognitive load. Awareness tools constitute as a kind of MM prosthesis,
        and, like any prosthesis, they many augment MM (by facilitating it or
        even scaffolding it) or inhibit it (by making it useless).

@@

Our main issue is the symmetry questions: what is the relationship
(Fig.~\ref{mm_symmetry}) between \Model{A}{B}{X} and
\Model{B}{A}{X}? A low symmetry would mean that mutual modeling is
mainly an individual attitude/aptitude ($\mathcal{H}_{1}$). A high correlation
might support $\mathcal{H}_{2}$ since there is a low probability that randomly
formed pairs integrate peers with the same level of MM skills.

@@

\caption{The MM symmetry question $\Delta_1 =  \Delta(\mathcal{M}^{\circ} (A,B,X),
\mathcal{M}^{\circ} (B,A,X))$}

@@

    \end{tikzpicture}
}
\caption{The MM triangle questions}

>>>>

In order to build a shared understanding, do partners have to build a
representation of each other's knowledge? We refer to \emph{mutual modeling} as
the process of inferring one's partner mental states. Any claim that students
carry out a detailed monitoring of their peers would be as incorrect as any
claim that they do not maintain any representation at all. If mutual modeling
had to be permanently detailed and accurate, subjects would obviously face a
huge cognitive load. Conversely, peers could not collaborate without some
minimal amount of mutual modeling. For instance, A cannot disagree with B
without knowing that B has a different opinion. The mutual model can be implicit
(A is not aware of what he knows about B), by-default (I believe that B beliefs
what I believe unless contrary evidence), opportunistic (A does not model B
unless the conversation requires it), global (A infers B's beliefs based on
categories such as age, culture or profession) and, of course, it can be
incorrect… but it can not remain empty. Dialogues include many instances of
utterances such as "I thought he would do that" (first level of mutual
modelling) or even "He thought I would do that but I intended something else."
(second level of mutual modelling).

@@

We have different levels of mutual modelling. If A states "B thinks I am god in maths", we
are the second level of mutual modelling:  $Model(A, B, Model(B,A, knowledge="good in
math"))$. There is an infinite regress of nested models: If A says "B knows that
I don't expect him to solve this statistics problem" corresponds to $Model(A, B,
Model (B, A (Model (A,B, statistic-skills)))$.

A partner model is probably not a "box", \ie not a monolithical representation
but rather a mosaic of information fragments about the partner, with various
grain sizes and various life cycles. This mosaic is elaborated through various
mechanisms, first for building an initial model of the partner and then for
updating this model.  As two students meet for the first time, mutual models are
initialized by the assumptions they make upon each other from cues such as
his/her membership to large categories (age, culture, profession, ...) include
stereotypes (sportsmen, junkie, business women, Swiss,...) as well as physical
appearance. Scholars studied how initial modeling impacts communication. In
their experiments on initial mutual modelling, Slugoski, Lalljee, Lamb \&
Ginsburg~\cite{slugoski1993attribution} pretended to their subjects that their
(fake) partner had or had not received the same information. They observed that
the subjects adapted their dialogue by focusing the explanation on the items
that (s)he was supposed to ignore. Brennan (1991) showed that the subjects used
different initial strategies in forming queries depending on who they were told
their partner was.

@@

Mutual modeling does not occur in a vacuum but it is highly contextualized. REF
reviews the features of the collaborative situation, namely the media
(cotemporality… ), may facilitate or hamper mutual modeling. Hutchins (SILENCE)
reported a study in which a short silence between two pilots was perfectly
interpreted because it occurred in a highly constrained communication context.
Some environments are more productive than others in helping peers to detect
their misunderstandings. Roschelle~\cite{roschelle1995construction} reformulate
the design of CSCL interfaces as providing ways for peer to detect and repairs
their misunderstanding. In CSCW, researchers have explored various so-called
"awareness tools", \ie functionalities that inform A is about B's actions that A
could not directly perceived because B was working on a different subset of the
virtual space. Different awareness tools will be addressed in this contribution
since they allowed us to manipulate experimentally the mutual modelling activity.

@@

\subsection{Related research of the applicants}

Pierre Dillenbourg has been working since several years on the role and importance
of mutual modelling in human-human interaction, specifically in the context of
learning sciences.

He already conducted several experiments~\cite{sangin2007partner,
nova2007collaboration} that supported the idea that modelling partners is not only
an individual skill. When engaged in a social task that could benefit from
mutual modelling, the accuracy of the models build by each of the participant on
the other participants increase.

In the frame of this work, Pierre Dillenbourg has proposed a simple formalism
that we describe below, and that we will use subsequently to describe the mutual
modelling situations.

\paragraph{A simple notational system for mutual modeling} We suggest
the notation \model{A}{B}{X} to denote ``A knows that B knows X". This is a
simple but hopefully useful reduction to reason on mutual modeling. This
notation does not mean A has an explicit, monolithic representation of B: it
must be understood as an abstraction referring to complex socio-cognitive
processes. Besides, we refer to the \emph{degree of accuracy} of the model as
\Model{A}{B}{X}.

To estimate the mutual modelling effort, we need 3 variables:

@@

    \item Tasks vary a lot with respect to how much they require.  The grounding
        criterion -- denoted $\mathcal{M}^{\circ}_{min}$ -- refers to how
        important it is to mutually share a piece of information X to succeed
        the task T. It can be computed as the probability to succeed T despite
        the fact X is not grounded. $\mathcal{M}^{\circ}_{min}(A,B,X)$ can be
        estimated from the correlation between \Model{A}{B}{X} and task
        performance.

    \item Before any specific grounding act, there is usually a non-null probability
        that X is mutually understood by A and B: because X is part of A's and
        B's cultures, because it is manifest to co-present subjects or simply
        because there is not much space for misunderstanding or disagreement
        about X. We could not simply collaborate without a high level of initial
        grounds. We note the theoretical accuracy of initial grounds:
        $\mathcal{M}^{\circ}_{t_0}(A,B,X)$.

    \item The cost of grounding X refers to the physical and cognitive effort
        necessary to perform a grounding act $\alpha$: a verbal repair (\eg
        rephrasing), a deictic gesture, a physical move to adopt one partner's
        viewpoint, etc. This cost varies according to media features ( Clark \&
        Brennan (REF)).

@@

Based on these 3 parameters, the probability of making an action $\alpha_{t_1}$ about
content X at time $t_1$ during task T in order to increase \Model{A}{B}{X}
at time $t_2$ is the ratio between how much it is needed  ((2)-(1)) and how much it
costs (3)~\cite{traum1996miscommunication}:

\begin{equation}
p(\alpha_{t_1}(X,T) / M^{\circ}_{t_2}(A,B,X) \nearrow) \approx \frac{(M^{\circ}_{min}(A,B,X,T) -
M^{\circ}_{t_1}(A,B,X))}{cost (\alpha)}
\end{equation}

This is qualitative summary more that a real equation since several parameters
are hard to quantify (\eg the cost an communication acts depends upon the user
as well) but it clarifies the parameters of the experiments we propose hereafter.

@@

    \item $\mathcal{M}^{\circ}_{t1}(A,B,X)$ Along the same reasoning, the initial degree
        of common grounds should be rather low (and hence the difference between
        initial and required degrees rather high) in order to make mutual
        modelling effort more observable. Studies 1, 4 and 5 have been conducted
        with teams of students who did not know each other. They came
        nonetheless from the same university: they hence had some general common
        grounds.  For studies 2 and 3, students knew each other before for
        reasons explained later on. In study 5, we manipulated the initial mutual modelling by
        using a JIGSAW script.

    \item $cost(\alpha)$ In all studies but study 4, the cost of  grounding is
        an independent variable. Study 3 uses media richness as independent
        variable, with the hypothesis that modeling emotions is "cheaper" with a
        richer medium, \ie  when peers can see each other.  Studies 1,2 and 4
        use awareness tools which, in principle, reduce the cost of mutual modelling, but do
        not eliminate all costs: if the tool provide A with information about
        what B does/knows, this additional information may actually increase
        cognitive load. Awareness tools constitute as a kind of mutual modelling prosthesis,
        and, like any prosthesis, they many augment mutual modelling (by facilitating it or
        even scaffolding it) or inhibit it (by making it useless).

@@

Our main issue is the symmetry questions: what is the relationship
(Fig.~\ref{mm_symmetry}) between \Model{A}{B}{X} and
\Model{B}{A}{X}? A low symmetry would mean that mutual modeling is
mainly an individual attitude/aptitude ($\mathcal{H}_{1}$). A high correlation
might support $\mathcal{H}_{2}$ since there is a low probability that randomly
formed pairs integrate peers with the same level of mutual modelling skills.

@@

\caption{The mutual modelling symmetry question $\Delta_1 =  \Delta(\mathcal{M}^{\circ} (A,B,X),
\mathcal{M}^{\circ} (B,A,X))$}

@@

    \end{tikzpicture}
}
\caption{The mutual modelling triangle questions}
